---
title: "DESeq2 Pipeline – HTN vs Control (Pooled & Chamber-Stratified)"
author: "Destiny Pounds & Micaiah McDonald"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 6
)
# Final analysis was ran on Destiny's count data
# Set different wd for future studies
setwd("/Users/destinypounds/Desktop/MSBD501/Data") 
```

##################################
# 0) Setup: install/load packages
##################################

```{r install-packages}
# Install BiocManager if not already present — required to install
# Bioconductor packages (DESeq2, apeglm, annotation databases)
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

# Bioconductor packages:
#   DESeq2       — core differential expression framework
#   apeglm       — adaptive shrinkage estimator for log2 fold changes
#   AnnotationDbi— interface for Bioconductor annotation databases
#   org.Hs.eg.db — human gene annotation (Ensembl → symbol/name)
#   limma        — used here only for removeBatchEffect() on VST matrix
bioc_pkgs <- c("DESeq2", "apeglm", "AnnotationDbi", "org.Hs.eg.db", "limma")

# CRAN packages:
#   ggplot2  — publication-quality plotting
#   pheatmap — clustered heatmaps with annotation bars
#   ggrepel  — non-overlapping text labels on ggplot2 figures
#   dplyr    — data manipulation utilities
cran_pkgs <- c("ggplot2", "pheatmap", "ggrepel", "dplyr")

for (pkg in bioc_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, ask = FALSE, update = FALSE)
  }
}
for (pkg in cran_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}
```

```{r load-packages}
library(DESeq2)
library(apeglm)
library(limma)        
library(AnnotationDbi)
library(org.Hs.eg.db)
library(ggplot2)
library(pheatmap)
library(ggrepel)
library(dplyr)
```

############################################################
# Step 1: Create output folder for all saved plots
############################################################

```{r output-dir}
# All figures (boxplots, PCAs, heatmaps, MA/volcano plots) are written
# to a single 'plots/' subfolder inside the working directory.
# dir.create() with showWarnings = FALSE is safe to re-run — it will
# not error if the folder already exists.
PLOT_DIR <- "plots"
dir.create(PLOT_DIR, showWarnings = FALSE)
cat("Plots will be saved to:", normalizePath(PLOT_DIR), "\n")
```

############################################################
# Step 2: Load the featureCounts count matrix
############################################################

```{r load-counts}
#Load featurecounts
fc <- read.delim(
  "/Users/destinypounds/Desktop/MSBD501/Data/featurecounts_counts.txt",
  comment.char     = "#",
  stringsAsFactors = FALSE,
  check.names      = FALSE
)

# Separate annotation from count data
gene_annot       <- fc[, 1:6]          # columns 1–6: gene metadata
counts           <- fc[, 7:ncol(fc)]   # columns 7+: raw integer counts
rownames(counts) <- fc$Geneid

# featureCounts prefixes each column name with the BAM path used.
# Strip "bam/" prefix and ".bam" suffix to get clean SRR accession IDs.
colnames(counts) <- gsub("^bam/",   "", colnames(counts))
colnames(counts) <- gsub("\\.bam$", "", colnames(counts))

# DESeq2 requires the count matrix to be an integer matrix.
counts <- as.matrix(counts)
storage.mode(counts) <- "integer"

#Check counts
head(counts[, 1:4])
```

############################################################
# Step 3: Sample metadata
############################################################

```{r sample-metadata}
# Columns:
#   condition — experimental group (Control vs HTN); Control is reference
#   ventricle — chamber of origin (Left vs Right); used as covariate
#   donor     — donor ID; paired LV/RV samples share the same donor number
#   label     — readable label used for visualization purposes

coldata <- data.frame(
  row.names = colnames(counts),
  condition = c(
    "Control", "Control", "HTN",     "HTN",     "Control",
    "Control", "HTN",     "HTN",     "HTN",     "HTN"
  ),
  ventricle = c(
    "Right",   "Left",    "Right",   "Left",    "Right",
    "Left",    "Right",   "Left",    "Right",   "Left"
  ),
  donor = factor(c(7, 7, 5, 5, 4, 4, 3, 3, 1, 1)),
  label = c(
    "Donor 7, RV", "Donor 7, LV",
    "Donor 5, RV", "Donor 5, LV",
    "Donor 4, RV", "Donor 4, LV",
    "Donor 3, RV", "Donor 3, LV",
    "Donor 1, RV", "Donor 1, LV"
  )
)

# Set factor levels so DESeq2 uses the correct reference:
#   condition: "Control" is ref 
#   ventricle: "Left"    is ref 
coldata$condition <- factor(coldata$condition, levels = c("Control", "HTN"))
coldata$ventricle <- factor(coldata$ventricle, levels = c("Left", "Right"))

#View column data
coldata
```

############################################################
# Step 4: Filter low-count genes (shared across all analyses)
############################################################

```{r filter-counts}
# Genes with very low counts carry little statistical information and
# inflate multiple-testing burden. We retain only genes with >= 10
# counts in at least 2 samples (the size of the smallest group).
keep   <- rowSums(counts >= 10) >= 2
counts <- counts[keep, ]
cat("Genes retained after filtering:", nrow(counts), "\n")
```

############################################################
# Helper functions
# Defined once — called identically across all 3 analyses.
############################################################

```{r helpers}
# ---------------------------------------------------------------------------
# annotate_results()
# Maps Ensembl gene IDs (with version suffix, e.g. ENSG00000141510.18)
# to human-readable gene symbols (e.g. TP53) and full gene names.
# Version suffixes are stripped with sub() before mapping.
# ---------------------------------------------------------------------------
annotate_results <- function(res_df) {
  res_df$gene_id    <- rownames(res_df)
  res_df$ensembl_id <- sub("\\..*$", "", res_df$gene_id)   # strip version

  res_df$gene_symbol <- mapIds(
    org.Hs.eg.db,
    keys      = res_df$ensembl_id,
    column    = "SYMBOL",
    keytype   = "ENSEMBL",
    multiVals = "first"
  )
  res_df$gene_name <- mapIds(
    org.Hs.eg.db,
    keys      = res_df$ensembl_id,
    column    = "GENENAME",
    keytype   = "ENSEMBL",
    multiVals = "first"
  )
  res_df
}

# ---------------------------------------------------------------------------
# mark_significance()
# Classifies each gene as Upregulated, Downregulated, or Not Significant
# using standard thresholds: padj < 0.05 AND |log2FoldChange| > 1.
#
# Also builds a balanced top-label set for plots: the n_each most
# significant upregulated genes + n_each most significant downregulated
# genes (ranked by padj). Balanced selection ensures both directions are
# represented in volcano and MA plot labels rather than one side dominating.
#
# Returns a named list:
#   $res_df     — full results with significance column 
#   $sig_only   — subset of significant genes only
#   $top_labels — balanced label subset for plot annotation
# ---------------------------------------------------------------------------
mark_significance <- function(res_df, padj_cut = 0.05, lfc_cut = 1, n_each = 5) {
  res_df$significance <- "Not Significant"
  res_df$significance[
    !is.na(res_df$padj) & res_df$padj < padj_cut & res_df$log2FoldChange >  lfc_cut
  ] <- "Upregulated"
  res_df$significance[
    !is.na(res_df$padj) & res_df$padj < padj_cut & res_df$log2FoldChange < -lfc_cut
  ] <- "Downregulated"

  # log10(baseMean + 1) used as x-axis in MA plots
  res_df$baseMean_log10 <- log10(res_df$baseMean + 1)

  sig_only <- subset(res_df, !is.na(padj) & padj < padj_cut & abs(log2FoldChange) > lfc_cut)

  # Split by direction, sort by padj, take top n_each from each side
  up_sig   <- sig_only[sig_only$log2FoldChange >  lfc_cut, ]
  down_sig <- sig_only[sig_only$log2FoldChange < -lfc_cut, ]

  top_labels <- rbind(
    head(up_sig[order(up_sig$padj), ],     n_each),
    head(down_sig[order(down_sig$padj), ], n_each)
  )

  # Fall back to Ensembl ID if gene symbol is missing
  top_labels$label_name <- ifelse(
    is.na(top_labels$gene_symbol) | top_labels$gene_symbol == "",
    rownames(top_labels),
    top_labels$gene_symbol
  )

  # Drop any labels with non-finite coordinates (would crash ggrepel)
  top_labels <- subset(top_labels, is.finite(baseMean_log10) & is.finite(log2FoldChange))

  list(res_df = res_df, sig_only = sig_only, top_labels = top_labels)
}

# ---------------------------------------------------------------------------
# make_ma_plot()
# Creates MA plot: log10(mean expression) on x-axis vs shrunken LFC on y-axis.
# ---------------------------------------------------------------------------
make_ma_plot <- function(res_df, top_labels, title) {
  p <- ggplot(res_df, aes(x = baseMean_log10, y = log2FoldChange, color = significance)) +
    geom_point(alpha = 0.6) +
    geom_hline(yintercept = c(-1, 1), linetype = "dashed") +
    scale_color_manual(values = c(
      "Upregulated"     = "firebrick",
      "Downregulated"   = "steelblue",
      "Not Significant" = "grey60"
    )) +
    xlab("log10(Base Mean + 1)") +
    ylab("Shrunken log2 Fold Change") +
    ggtitle(title) +
    theme_minimal()

  if (nrow(top_labels) > 0) {
    p <- p + ggrepel::geom_text_repel(
      data        = top_labels,
      inherit.aes = FALSE,
      aes(x = baseMean_log10, y = log2FoldChange, label = label_name),
      size = 3.5, max.overlaps = 20
    )
  }
  p
}

# ---------------------------------------------------------------------------
# make_volcano_plot()
# Creates Volcano plot: LFC on x-axis vs -log10(padj) on y-axis.
# ---------------------------------------------------------------------------
make_volcano_plot <- function(res_df, top_labels, title, lfc_cut = 1) {
  p <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
    geom_point(alpha = 0.6) +
    geom_vline(xintercept = c(-lfc_cut, lfc_cut), linetype = "dashed") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    scale_color_manual(values = c(
      "Upregulated"     = "firebrick",
      "Downregulated"   = "steelblue",
      "Not Significant" = "grey60"
    )) +
    xlab("log2 Fold Change") +
    ylab("-log10 Adjusted P-value") +
    ggtitle(title) +
    theme_minimal()

  if (nrow(top_labels) > 0) {
    p <- p + ggrepel::geom_text_repel(
      data        = top_labels,
      aes(label = label_name),
      size = 3.5, max.overlaps = 20, show.legend = FALSE
    )
  }
  p
}

# ---------------------------------------------------------------------------
# make_heatmap()
# Creates heatmap of the top n_genes significant DE genes (row-scaled z-scores).
# Columns are ordered manually (Control before HTN, LV before RV within
# each condition) rather than clustered, so biological grouping is visible.
# Donor labels replace SRR accession numbers on column headers.
# annonation_vars controls what labels are included in the legend.
# ---------------------------------------------------------------------------
make_heatmap <- function(vsd_obj, sig_only, coldata_sub,
                         annotation_vars = c("condition", "ventricle"),
                         label_col       = "label",
                         order_by        = c("condition", "ventricle"),
                         title           = "Top DE Genes",
                         n_genes         = 30) {
  if (nrow(sig_only) == 0) {
    message("No significant genes — heatmap skipped.")
    return(invisible(NULL))
  }

  # Select top n genes (already ordered by padj in sig_only)
  top_genes  <- rownames(sig_only)[1:min(n_genes, nrow(sig_only))]
  mat        <- assay(vsd_obj)[top_genes, ]

  # Row-wise z-score scaling so expression patterns are comparable
  # across genes with different absolute expression levels
  mat_scaled <- t(scale(t(mat)))

  # Use gene symbol as row label; fall back to Ensembl ID if symbol is NA
  row_lbl <- sig_only[match(top_genes, rownames(sig_only)), "gene_symbol"]
  row_lbl[is.na(row_lbl)] <- top_genes[is.na(row_lbl)]
  rownames(mat_scaled) <- row_lbl

  # Reorder columns: e.g. by condition then ventricle
  col_order      <- do.call(order, lapply(order_by, function(v) coldata_sub[[v]]))
  mat_scaled     <- mat_scaled[, col_order]
  ann_clean      <- coldata_sub[col_order, annotation_vars, drop = FALSE]

  # Replace SRR column names with donor labels
  colnames(mat_scaled) <- coldata_sub[[label_col]][col_order]
  rownames(ann_clean)  <- coldata_sub[[label_col]][col_order]

  pheatmap(
    mat_scaled,
    annotation_col = ann_clean,
    show_rownames  = TRUE,
    cluster_cols   = FALSE,   # respect manual column ordering
    cluster_rows   = TRUE,    # hierarchically cluster genes
    fontsize        = 12,
    fontsize_row    = 11,
    fontsize_col    = 11,
    main            = title
  )
}
```

############################################################
# ══════════════════════════════════════════════════════════
#   ANALYSIS 1 – POOLED  (~ ventricle + condition)
#
#   All 10 samples are used together. Ventricle (LV vs RV)
#   is included as covariate so its expression
#   effect is removed.
# ══════════════════════════════════════════════════════════
############################################################

## 1.1  Build DESeq2 object – pooled

```{r deseq-pooled}
# Design: ~ ventricle + condition
#   ventricle — covariate; adjusts for LV/RV expression differences
#   condition — term of interest; HTN_vs_Control coefficient is extracted
# DESeq() runs: estimation of size factors → dispersion → negative
# binomial GLM → Wald test for each gene
dds_pooled <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = coldata,
  design    = ~ ventricle + condition
)
dds_pooled <- DESeq(dds_pooled)
resultsNames(dds_pooled)   # confirm coefficient names before extracting results

normalized_counts_pooled <- counts(dds_pooled, normalized = TRUE)
```

## 1.2  Raw vs normalized count boxplots – pooled

```{r boxplots-pooled}
# Boxplots on log2 scale give a quick visual check that:
#   Raw:        library sizes differ across samples (unequal medians expected)
#   Normalized: size-factor correction brings medians into alignment
# Persistent spread differences after normalization may indicate batch effects
# or outlier samples worth investigating before proceeding.

# --- Raw counts ---
png(file.path(PLOT_DIR, "boxplot_raw_counts.png"),
    width = 1400, height = 800, res = 150)
par(mar = c(8, 4, 4, 2))
boxplot(
  log2(counts + 1),
  names    = coldata$label,
  main     = "Raw Gene Expression Counts (All Samples)",
  ylab     = "log2(count + 1)",
  las      = 2,
  cex.axis = 0.8,
  col      = "grey80"
)
dev.off()

# Repeat plotting for inline viewing
par(mar = c(8, 4, 4, 2))
boxplot(
  log2(counts + 1),
  names    = coldata$label,
  main     = "Raw Gene Expression Counts (All Samples)",
  ylab     = "log2(count + 1)",
  las      = 2,
  cex.axis = 0.8,
  col      = "grey80"
)

# --- Normalized counts ---
png(file.path(PLOT_DIR, "boxplot_normalized_counts.png"),
    width = 1400, height = 800, res = 150)
par(mar = c(8, 4, 4, 2))
boxplot(
  log2(normalized_counts_pooled + 1),
  names    = coldata$label,
  main     = "Normalized Gene Expression – Pooled Model",
  ylab     = "log2(normalized count + 1)",
  las      = 2,
  cex.axis = 0.8,
  col      = "grey80"
)
dev.off()

#Repeat plotting for inline viewing
par(mar = c(8, 4, 4, 2))
boxplot(
  log2(normalized_counts_pooled + 1),
  names    = coldata$label,
  main     = "Normalized Gene Expression – Pooled Model",
  ylab     = "log2(normalized count + 1)",
  las      = 2,
  cex.axis = 0.8,
  col      = "grey80"
)

```

## 1.3  VST & PCA – pooled (ventricle batch-corrected for visualization only)

```{r pca-pooled}
# VST (variance stabilizing transformation) converts raw counts to a
# roughly homoskedastic log-scale matrix suitable for distance-based
# visualization. blind = FALSE uses the design to inform the dispersion
# fit — recommended when the design is well-specified.
vsd_pooled <- vst(dds_pooled, blind = FALSE)

# removeBatchEffect() from limma regresses out the ventricle effect
# from the VST matrix FOR VISUALIZATION ONLY. This makes condition
# separation visible in the PCA without ventricle dominating PC1.
# IMPORTANT: this modified assay is NOT fed back into DESeq2 — the
# actual DE model already accounts for ventricle via the design formula.
assay(vsd_pooled) <- removeBatchEffect(
  assay(vsd_pooled),
  batch = colData(vsd_pooled)$ventricle
)

pca_data   <- plotPCA(vsd_pooled, intgroup = c("condition", "ventricle"), returnData = TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))

# Replace SRR row names with donor labels for readable plot text
pca_data$label <- coldata[rownames(pca_data), "label"]

pca_pooled <- ggplot(pca_data,
    aes(PC1, PC2, color = condition, shape = ventricle, label = label)) +
  geom_point(size = 4) +
  geom_text_repel(size = 3.7, max.overlaps = 20) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA – Pooled Analysis (HTN vs Control, adjusted for chamber)") +
  theme_minimal()

print(pca_pooled)
ggsave(file.path(PLOT_DIR, "pca_pooled.png"),
       plot = pca_pooled, width = 8, height = 6, dpi = 150)
```

## 1.3b  PCA on DEGs only – pooled

```{r pca-pooled-degs}
# A second PCA restricted to statistically significant DEGs (padj < 0.05,
# no LFC threshold here) reduces noise from the ~20,000 non-DE genes and
# can reveal cleaner separation. 
vsd_fresh <- vst(dds_pooled, blind = FALSE)

res_tmp   <- as.data.frame(results(dds_pooled, contrast = c("condition", "HTN", "Control")))
sig_genes <- rownames(res_tmp[!is.na(res_tmp$padj) & res_tmp$padj < 0.05, ])
cat("DEGs used for PCA:", length(sig_genes), "\n")

# Correct only the DEG rows for ventricle before running PCA
mat_sig_corrected <- removeBatchEffect(
  assay(vsd_fresh)[sig_genes, ],
  batch = colData(vsd_fresh)$ventricle
)

# prcomp() with scale. = TRUE standardizes variance across genes,
# preventing high-variance genes from dominating the components
pca_res    <- prcomp(t(mat_sig_corrected), scale. = TRUE)
percentVar <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2))

pca_deg_df <- data.frame(
  PC1       = pca_res$x[, 1],
  PC2       = pca_res$x[, 2],
  condition = colData(vsd_fresh)$condition,
  ventricle = colData(vsd_fresh)$ventricle,
  label     = coldata$label
)

pca_degs <- ggplot(pca_deg_df,
    aes(PC1, PC2, color = condition, shape = ventricle, label = label)) +
  geom_point(size = 4) +
  geom_text_repel(size = 3.7, max.overlaps = 20) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA – DEGs Only (HTN vs Control, adjusted for ventricle)") +
  theme_minimal()

print(pca_degs)
ggsave(file.path(PLOT_DIR, "pca_pooled_degs.png"),
       plot = pca_degs, width = 8, height = 6, dpi = 150)
```

## 1.4  Sample-to-sample distance heatmap – pooled

```{r dist-heatmap-pooled}
# Euclidean distances between samples in VST space reveal global
# clustering structure. Samples from the same condition or chamber
# that cluster together confirm expected biological groupings.
sampleDistMatrix  <- as.matrix(dist(t(assay(vsd_pooled))))

dist_col_order_pooled <- order(coldata$condition, coldata$ventricle)
dist_labels_pooled    <- coldata$label[dist_col_order_pooled]

sampleDistMatrix_ord           <- sampleDistMatrix[dist_col_order_pooled, dist_col_order_pooled]
colnames(sampleDistMatrix_ord) <- dist_labels_pooled
rownames(sampleDistMatrix_ord) <- dist_labels_pooled

annotation_pooled_ord           <- coldata[dist_col_order_pooled, c("condition", "ventricle"), drop = FALSE]
rownames(annotation_pooled_ord) <- dist_labels_pooled

dist_heatmap_pooled <- pheatmap(
  sampleDistMatrix_ord,
  annotation_col = annotation_pooled_ord,
  labels_col     = dist_labels_pooled,
  labels_row     = dist_labels_pooled,
  cluster_rows   = FALSE,
  cluster_cols   = FALSE,
  fontsize        = 12,
  main            = "Sample-to-sample Distance – Pooled"
)
print(dist_heatmap_pooled)
ggsave(file.path(PLOT_DIR, "distance_heatmap_pooled.png"),
       plot = dist_heatmap_pooled, width = 8, height = 6, dpi = 150)
```

## 1.5  DE results & LFC shrinkage – pooled

```{r results-pooled}
# Extract the HTN vs Control contrast from the pooled model.
res_pooled <- results(dds_pooled, contrast = c("condition", "HTN", "Control"))
res_pooled <- res_pooled[order(res_pooled$padj), ]
summary(res_pooled)

# lfcShrink() with type = "apeglm" applies adaptive shrinkage to
# log2FoldChange estimates. This pulls noisy LFC estimates from low-count
# genes toward zero, producing more reliable fold-change rankings.
res_pooled_shrunk <- lfcShrink(dds_pooled,
  coef = "condition_HTN_vs_Control", type = "apeglm")
res_pooled_shrunk <- res_pooled_shrunk[order(res_pooled_shrunk$padj), ]
res_df_pooled     <- as.data.frame(res_pooled_shrunk)
```

## 1.6  Annotate – pooled

```{r annotate-pooled}
# Map Ensembl IDs to gene symbols and full names via org.Hs.eg.db.
res_df_pooled <- annotate_results(res_df_pooled)
```

## 1.7  Significance & top labels – pooled

```{r sig-pooled}
# Thresholds: padj < 0.05 (FDR) AND |LFC| > 1
# (at least 2-fold change after shrinkage). Both criteria must be met.
# Top labels: 5 most significant upregulated + 5 most significant
# downregulated genes (by padj) for balanced MA and volcano annotations.
pooled_out        <- mark_significance(res_df_pooled)
res_df_pooled     <- pooled_out$res_df
sig_only_pooled   <- pooled_out$sig_only
top_labels_pooled <- pooled_out$top_labels

table(res_df_pooled$significance)
cat("Significant DEGs (pooled):", nrow(sig_only_pooled), "\n")
head(sig_only_pooled[, c("gene_symbol", "gene_name", "log2FoldChange", "padj")])
```

## 1.8  MA plot – pooled

```{r ma-pooled}
p_ma_pooled <- make_ma_plot(res_df_pooled, top_labels_pooled,
  "MA Plot – Pooled (HTN vs Control, adjusted for chamber)")
print(p_ma_pooled)
ggsave(file.path(PLOT_DIR, "ma_plot_pooled.png"),
       plot = p_ma_pooled, width = 8, height = 6, dpi = 150)
```

## 1.9  Volcano plot – pooled

```{r volcano-pooled}
# Volcano plot displays statistical significance (-log10 padj) against
# effect size (log2FC) for every gene. Genes in the upper corners
# (true DEGs) are highlighted and labeled.
volcano_pooled <- make_volcano_plot(
  res_df_pooled, top_labels_pooled,
  "Volcano Plot – Pooled (HTN vs Control, adjusted for chamber)"
)
print(volcano_pooled)
ggsave(file.path(PLOT_DIR, "volcano_pooled.png"),
       plot = volcano_pooled, width = 8, height = 6, dpi = 150)
```

## 1.10  Heatmap – pooled

```{r heatmap-pooled}
# Heatmap of top 30 significant DE genes.
top_heatmap_pooled <- make_heatmap(
  vsd_obj     = vsd_pooled,
  sig_only    = sig_only_pooled,
  coldata_sub = coldata,
  title       = "Top DE Genes – Pooled Analysis"
)
if (!is.null(top_heatmap_pooled)) {
  print(top_heatmap_pooled)
  ggsave(file.path(PLOT_DIR, "heatmap_pooled.png"),
         plot = top_heatmap_pooled, width = 10, height = 8, dpi = 150)
}
```

## 1.11  Top 20 significant genes – pooled

```{r top20-pooled}
# Preview of the 20 most significant DEGs for a quick check
top20_pooled <- head(sig_only_pooled[order(sig_only_pooled$padj), ], 20)
top20_pooled[, c("gene_symbol", "gene_name", "log2FoldChange", "padj")]
```

## 1.12  Save pooled output

```{r save-pooled}
# Four CSV files per analysis:
#   *_full.csv        — all genes with shrunken LFC, padj, and annotation
#   *_significant.csv — significant genes only (padj < 0.05 & |LFC| > 1)
#   *_top20.csv       — top 20 by padj for quick reference
#   normalized_counts — size-factor-corrected count matrix
write.csv(res_df_pooled,            "DESeq2_Pooled_HTN_vs_Control.csv",            row.names = FALSE)
write.csv(sig_only_pooled,          "DESeq2_Pooled_HTN_vs_Control_significant.csv", row.names = FALSE)
write.csv(top20_pooled,             "DESeq2_Pooled_top20.csv",                      row.names = FALSE)
write.csv(normalized_counts_pooled, "normalized_counts_pooled.csv")
```

############################################################
# ══════════════════════════════════════════════════════════
#   ANALYSIS 2 – LV ONLY  (~ condition)
#
#   Subset to Left Ventricle samples only. With a single
#   chamber there is no ventricle covariate, so the design
#   simplifies to ~ condition. Fewer samples means reduced
#   power, but chamber-specific results can reveal LV-
#   selective HTN gene expression changes.
# ══════════════════════════════════════════════════════════
############################################################

## 2.1  Subset to LV

```{r subset-lv}
# Subset count matrix and metadata to LV samples only.
# droplevels() removes the unused "Right" level from the ventricle factor
# so DESeq2 does not try to estimate a coefficient for a missing group.
# A fresh per-subset low-count filter is applied because genes that pass
# the global filter may still be very lowly expressed in LV samples alone.
lv_samples           <- rownames(coldata)[coldata$ventricle == "Left"]
counts_lv            <- counts[, lv_samples]
coldata_lv           <- coldata[lv_samples, , drop = FALSE]
coldata_lv$ventricle <- droplevels(coldata_lv$ventricle)

keep_lv   <- rowSums(counts_lv >= 10) >= 2
counts_lv <- counts_lv[keep_lv, ]
cat("LV samples:", ncol(counts_lv), "| Genes retained:", nrow(counts_lv), "\n")
coldata_lv
```

## 2.2  DESeq2 – LV only

```{r deseq-lv}
dds_lv <- DESeqDataSetFromMatrix(
  countData = counts_lv, colData = coldata_lv, design = ~ condition)
dds_lv <- DESeq(dds_lv)
resultsNames(dds_lv)
normalized_counts_lv <- counts(dds_lv, normalized = TRUE)
```

## 2.3  Raw vs normalized count boxplots – LV only

```{r boxplots-lv}
# LV-only boxplots confirm normalization within this subset.
#Raw plot
png(file.path(PLOT_DIR, "boxplot_raw_counts_lv.png"),
    width = 1000, height = 700, res = 150)
par(mar = c(8, 4, 4, 2))
boxplot(
  log2(counts_lv + 1),
  names    = coldata_lv$label,
  main     = "Raw Gene Expression Counts – LV Only",
  ylab     = "log2(count + 1)",
  las      = 2, cex.axis = 0.8,
  col      = "grey80"
)
dev.off()

#Plot again
par(mar = c(8, 4, 4, 2))
boxplot(
  log2(counts_lv + 1),
  names    = coldata_lv$label,
  main     = "Raw Gene Expression Counts – LV Only",
  ylab     = "log2(count + 1)",
  las      = 2, cex.axis = 0.8,
  col      = "grey80"
)

#Normalized plot
png(file.path(PLOT_DIR, "boxplot_normalized_counts_lv.png"),
    width = 1000, height = 700, res = 150)
par(mar = c(8, 4, 4, 2))
boxplot(
  log2(normalized_counts_lv + 1),
  names    = coldata_lv$label,
  main     = "Normalized Gene Expression – LV Only",
  ylab     = "log2(normalized count + 1)",
  las      = 2, cex.axis = 0.8,
  col      = "grey80"
)

dev.off()
#Plot again
par(mar = c(8, 4, 4, 2))
boxplot(
  log2(normalized_counts_lv + 1),
  names    = coldata_lv$label,
  main     = "Normalized Gene Expression – LV Only",
  ylab     = "log2(normalized count + 1)",
  las      = 2, cex.axis = 0.8,
  col      = "grey80"
)

```

## 2.4  VST & PCA – LV only

```{r pca-lv}
# PCA Plot
vsd_lv        <- vst(dds_lv, blind = FALSE)
pca_data_lv   <- plotPCA(vsd_lv, intgroup = "condition", returnData = TRUE)
percentVar_lv <- round(100 * attr(pca_data_lv, "percentVar"))
pca_data_lv$label <- coldata_lv[rownames(pca_data_lv), "label"]

pca_lv <- ggplot(pca_data_lv, aes(PC1, PC2, color = condition, label = label)) +
  geom_point(size = 4) +
  geom_text_repel(size = 3.7, max.overlaps = 20) +
  xlab(paste0("PC1: ", percentVar_lv[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_lv[2], "% variance")) +
  ggtitle("PCA – Left Ventricle Only (HTN vs Control)") +
  theme_minimal()

print(pca_lv)
ggsave(file.path(PLOT_DIR, "pca_lv.png"),
       plot = pca_lv, width = 8, height = 6, dpi = 150)
```

## 2.5  Sample distance heatmap – LV only

```{r dist-heatmap-lv}
# Order columns/rows: Control before HTN
dist_col_order_lv             <- order(coldata_lv$condition)
dist_labels_lv                <- coldata_lv$label[dist_col_order_lv]

sampleDistMatrix_lv           <- as.matrix(dist(t(assay(vsd_lv))))
sampleDistMatrix_lv_ord       <- sampleDistMatrix_lv[dist_col_order_lv, dist_col_order_lv]
colnames(sampleDistMatrix_lv_ord) <- dist_labels_lv
rownames(sampleDistMatrix_lv_ord) <- dist_labels_lv

annotation_lv_ord             <- coldata_lv[dist_col_order_lv, "condition", drop = FALSE]
rownames(annotation_lv_ord)   <- dist_labels_lv

dist_heatmap_lv <- pheatmap(
  sampleDistMatrix_lv_ord,
  annotation_col = annotation_lv_ord,
  labels_col     = dist_labels_lv,
  labels_row     = dist_labels_lv,
  cluster_rows   = FALSE,
  cluster_cols   = FALSE,
  fontsize = 12,
  main = "Sample-to-sample Distance – LV Only"
)
print(dist_heatmap_lv)
ggsave(file.path(PLOT_DIR, "distance_heatmap_lv.png"),
       plot = dist_heatmap_lv, width = 8, height = 6, dpi = 150)
```

## 2.6  DE results & LFC shrinkage – LV only

```{r results-lv}
res_lv <- results(dds_lv, contrast = c("condition", "HTN", "Control"))
summary(res_lv[order(res_lv$padj), ])

res_lv_shrunk <- lfcShrink(dds_lv,
  coef = "condition_HTN_vs_Control", type = "apeglm")
res_df_lv <- as.data.frame(res_lv_shrunk[order(res_lv_shrunk$padj), ])
```

## 2.7  Annotate – LV only

```{r annotate-lv}
res_df_lv <- annotate_results(res_df_lv)
```

## 2.8  Significance & top labels – LV only

```{r sig-lv}
lv_out        <- mark_significance(res_df_lv)
res_df_lv     <- lv_out$res_df
sig_only_lv   <- lv_out$sig_only
top_labels_lv <- lv_out$top_labels

table(res_df_lv$significance)
cat("Significant DEGs (LV):", nrow(sig_only_lv), "\n")
```

## 2.9  MA plot – LV only

```{r ma-lv}
p_ma_lv <- make_ma_plot(res_df_lv, top_labels_lv,
  "MA Plot – Left Ventricle Only (HTN vs Control)")
print(p_ma_lv)
ggsave(file.path(PLOT_DIR, "ma_plot_lv.png"),
       plot = p_ma_lv, width = 8, height = 6, dpi = 150)
```

## 2.10  Volcano plot – LV only

```{r volcano-lv}
volcano_lv <- make_volcano_plot(res_df_lv, top_labels_lv,
  "Volcano Plot – Left Ventricle Only (HTN vs Control)")
print(volcano_lv)
ggsave(file.path(PLOT_DIR, "volcano_lv.png"),
       plot = volcano_lv, width = 8, height = 6, dpi = 150)
```

## 2.11  Heatmap – LV only

```{r heatmap-lv}
top_heatmap_lv <- make_heatmap(
  vsd_obj         = vsd_lv,
  sig_only        = sig_only_lv,
  coldata_sub     = coldata_lv,
  annotation_vars = "condition",
  order_by        = "condition",
  title           = "Top DE Genes – Left Ventricle Only"
)
if (!is.null(top_heatmap_lv)) {
  print(top_heatmap_lv)
  ggsave(file.path(PLOT_DIR, "heatmap_lv.png"),
         plot = top_heatmap_lv, width = 10, height = 8, dpi = 150)
}
```

## 2.12  Top 20 – LV only

```{r top20-lv}
top20_lv <- head(sig_only_lv[order(sig_only_lv$padj), ], 20)
top20_lv[, c("gene_symbol", "gene_name", "log2FoldChange", "padj")]
```

## 2.13  Save LV output

```{r save-lv}
write.csv(res_df_lv,           "DESeq2_LV_HTN_vs_Control.csv",            row.names = FALSE)
write.csv(sig_only_lv,         "DESeq2_LV_HTN_vs_Control_significant.csv", row.names = FALSE)
write.csv(top20_lv,            "DESeq2_LV_top20.csv",                      row.names = FALSE)
write.csv(normalized_counts_lv,"normalized_counts_lv.csv")
```

############################################################
# ══════════════════════════════════════════════════════════
#   ANALYSIS 3 – RV ONLY  (~ condition)
#
#   Mirror of Analysis 2 for the Right Ventricle. Comparing
#   LV-only and RV-only results against the pooled analysis
#   reveals chamber-specific vs shared HTN gene responses.
# ══════════════════════════════════════════════════════════
############################################################

## 3.1  Subset to RV

```{r subset-rv}
rv_samples           <- rownames(coldata)[coldata$ventricle == "Right"]
counts_rv            <- counts[, rv_samples]
coldata_rv           <- coldata[rv_samples, , drop = FALSE]
coldata_rv$ventricle <- droplevels(coldata_rv$ventricle)

keep_rv   <- rowSums(counts_rv >= 10) >= 2
counts_rv <- counts_rv[keep_rv, ]
cat("RV samples:", ncol(counts_rv), "| Genes retained:", nrow(counts_rv), "\n")
coldata_rv
```

## 3.2  DESeq2 – RV only

```{r deseq-rv}
dds_rv <- DESeqDataSetFromMatrix(
  countData = counts_rv, colData = coldata_rv, design = ~ condition)
dds_rv <- DESeq(dds_rv)
resultsNames(dds_rv)
normalized_counts_rv <- counts(dds_rv, normalized = TRUE)
```

## 3.3  Raw vs normalized count boxplots – RV only

```{r boxplots-rv}
#Raw plots
png(file.path(PLOT_DIR, "boxplot_raw_counts_rv.png"),
    width = 1000, height = 700, res = 150)
par(mar = c(8, 4, 4, 2))
boxplot(
  log2(counts_rv + 1),
  names    = coldata_rv$label,
  main     = "Raw Gene Expression Counts – RV Only",
  ylab     = "log2(count + 1)",
  las      = 2, cex.axis = 0.8,
  col      = "grey80"
)
dev.off()

par(mar = c(8, 4, 4, 2))
boxplot(
  log2(counts_rv + 1),
  names    = coldata_rv$label,
  main     = "Raw Gene Expression Counts – RV Only",
  ylab     = "log2(count + 1)",
  las      = 2, cex.axis = 0.8,
  col      = "grey80"
)

#Normalized plots
png(file.path(PLOT_DIR, "boxplot_normalized_counts_rv.png"),
    width = 1000, height = 700, res = 150)
par(mar = c(8, 4, 4, 2))
boxplot(
  log2(normalized_counts_rv + 1),
  names    = coldata_rv$label,
  main     = "Normalized Gene Expression – RV Only",
  ylab     = "log2(normalized count + 1)",
  las      = 2, cex.axis = 0.8,
  col      = "grey80"
)

dev.off()

par(mar = c(8, 4, 4, 2))
boxplot(
  log2(normalized_counts_rv + 1),
  names    = coldata_rv$label,
  main     = "Normalized Gene Expression – RV Only",
  ylab     = "log2(normalized count + 1)",
  las      = 2, cex.axis = 0.8,
  col      = "grey80"
)
```

## 3.4  VST & PCA – RV only

```{r pca-rv}
vsd_rv        <- vst(dds_rv, blind = FALSE)
pca_data_rv   <- plotPCA(vsd_rv, intgroup = "condition", returnData = TRUE)
percentVar_rv <- round(100 * attr(pca_data_rv, "percentVar"))
pca_data_rv$label <- coldata_rv[rownames(pca_data_rv), "label"]

pca_rv <- ggplot(pca_data_rv, aes(PC1, PC2, color = condition, label = label)) +
  geom_point(size = 4) +
  geom_text_repel(size = 3.7, max.overlaps = 20) +
  xlab(paste0("PC1: ", percentVar_rv[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_rv[2], "% variance")) +
  ggtitle("PCA – Right Ventricle Only (HTN vs Control)") +
  theme_minimal()

print(pca_rv)
ggsave(file.path(PLOT_DIR, "pca_rv.png"),
       plot = pca_rv, width = 8, height = 6, dpi = 150)
```

## 3.5  Sample distance heatmap – RV only

```{r dist-heatmap-rv}
# Order columns/rows: Control before HTN
dist_col_order_rv             <- order(coldata_rv$condition)
dist_labels_rv                <- coldata_rv$label[dist_col_order_rv]

sampleDistMatrix_rv           <- as.matrix(dist(t(assay(vsd_rv))))
sampleDistMatrix_rv_ord       <- sampleDistMatrix_rv[dist_col_order_rv, dist_col_order_rv]
colnames(sampleDistMatrix_rv_ord) <- dist_labels_rv
rownames(sampleDistMatrix_rv_ord) <- dist_labels_rv

annotation_rv_ord             <- coldata_rv[dist_col_order_rv, "condition", drop = FALSE]
rownames(annotation_rv_ord)   <- dist_labels_rv

dist_heatmap_rv <- pheatmap(
  sampleDistMatrix_rv_ord,
  annotation_col = annotation_rv_ord,
  labels_col     = dist_labels_rv,
  labels_row     = dist_labels_rv,
  cluster_rows   = FALSE,
  cluster_cols   = FALSE,
  fontsize = 12,
  main = "Sample-to-sample Distance – RV Only"
)
print(dist_heatmap_rv)
ggsave(file.path(PLOT_DIR, "distance_heatmap_rv.png"),
       plot = dist_heatmap_rv, width = 8, height = 6, dpi = 150)
```

## 3.6  DE results & LFC shrinkage – RV only

```{r results-rv}
res_rv <- results(dds_rv, contrast = c("condition", "HTN", "Control"))
summary(res_rv[order(res_rv$padj), ])

res_rv_shrunk <- lfcShrink(dds_rv,
  coef = "condition_HTN_vs_Control", type = "apeglm")
res_df_rv <- as.data.frame(res_rv_shrunk[order(res_rv_shrunk$padj), ])
```

## 3.7  Annotate – RV only

```{r annotate-rv}
res_df_rv <- annotate_results(res_df_rv)
```

## 3.8  Significance & top labels – RV only

```{r sig-rv}
rv_out        <- mark_significance(res_df_rv)
res_df_rv     <- rv_out$res_df
sig_only_rv   <- rv_out$sig_only
top_labels_rv <- rv_out$top_labels

table(res_df_rv$significance)
cat("Significant DEGs (RV):", nrow(sig_only_rv), "\n")
```

## 3.9  MA plot – RV only

```{r ma-rv}
p_ma_rv <- make_ma_plot(res_df_rv, top_labels_rv,
  "MA Plot – Right Ventricle Only (HTN vs Control)")
print(p_ma_rv)
ggsave(file.path(PLOT_DIR, "ma_plot_rv.png"),
       plot = p_ma_rv, width = 8, height = 6, dpi = 150)
```

## 3.10  Volcano plot – RV only

```{r volcano-rv}
volcano_rv <- make_volcano_plot(res_df_rv, top_labels_rv,
  "Volcano Plot – Right Ventricle Only (HTN vs Control)")
print(volcano_rv)
ggsave(file.path(PLOT_DIR, "volcano_rv.png"),
       plot = volcano_rv, width = 8, height = 6, dpi = 150)
```

## 3.11  Heatmap – RV only

```{r heatmap-rv}
top_heatmap_rv <- make_heatmap(
  vsd_obj         = vsd_rv,
  sig_only        = sig_only_rv,
  coldata_sub     = coldata_rv,
  annotation_vars = "condition",
  order_by        = "condition",
  title           = "Top DE Genes – Right Ventricle Only"
)
if (!is.null(top_heatmap_rv)) {
  print(top_heatmap_rv)
  ggsave(file.path(PLOT_DIR, "heatmap_rv.png"),
         plot = top_heatmap_rv, width = 10, height = 8, dpi = 150)
}
```

## 3.12  Top 20 – RV only

```{r top20-rv}
top20_rv <- head(sig_only_rv[order(sig_only_rv$padj), ], 20)
top20_rv[, c("gene_symbol", "gene_name", "log2FoldChange", "padj")]
```

## 3.13  Save RV output

```{r save-rv}
write.csv(res_df_rv,           "DESeq2_RV_HTN_vs_Control.csv",            row.names = FALSE)
write.csv(sig_only_rv,         "DESeq2_RV_HTN_vs_Control_significant.csv", row.names = FALSE)
write.csv(top20_rv,            "DESeq2_RV_top20.csv",                      row.names = FALSE)
write.csv(normalized_counts_rv,"normalized_counts_rv.csv")
```

############################################################
# ══════════════════════════════════════════════════════════
#   Overlap summary: DEGs shared across all three analyses
#
#   Genes significant in all three analyses are the most
#   robust HTN candidates — they survive both the full
#   pooled model and chamber-specific subsetting.
#   Genes unique to LV or RV suggest chamber-selective
#   transcriptional responses to hypertension.
# ══════════════════════════════════════════════════════════
############################################################

```{r overlap-summary}
sig_pooled_ids <- rownames(sig_only_pooled)
sig_lv_ids     <- rownames(sig_only_lv)
sig_rv_ids     <- rownames(sig_only_rv)

cat("Significant DEGs (padj < 0.05 & |LFC| > 1):\n")
cat("  Pooled (ventricle + condition):", length(sig_pooled_ids), "\n")
cat("  LV only:                       ", length(sig_lv_ids),     "\n")
cat("  RV only:                       ", length(sig_rv_ids),     "\n")
cat("  Pooled ∩ LV:                   ", length(intersect(sig_pooled_ids, sig_lv_ids)), "\n")
cat("  Pooled ∩ RV:                   ", length(intersect(sig_pooled_ids, sig_rv_ids)), "\n")
cat("  LV ∩ RV (shared chamber DEGs):", length(intersect(sig_lv_ids, sig_rv_ids)),     "\n")
cat("  All three:                     ", length(Reduce(intersect,
      list(sig_pooled_ids, sig_lv_ids, sig_rv_ids))), "\n")
```
